---
output:
  html_document: default
  pdf_document: default
---
TITLE by YOUR_NAME_HERE
========================================================

```{r global_options, include = FALSE}
knitr::opts_chunk$set(echo=FALSE) #set echo to FALSE for all code chunks
```


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(dplyr)
library(scales)
library(gender)
library(genderdata)
library(hash) #using for gender stuff
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
# data source: http://fec.gov/disclosurep/pnational.do
setwd('/Users/chadhorner/Documents/OneDrive/Udacity/P4')
newjersey = read.csv('newjersey.csv')
newjersey$X. = NULL #eliminate extra column
attach(newjersey) #make each column its own variable
```

```{r echo=FALSE, initial_examination}
names(newjersey)
# cand_id/cand_nm: id/name of candidate
# contbr_name: name of donor
# contbr_city: city of donor (could be interesting)
# contbr_zip: zip code of donor (can probably get certain data by zip)
# contbr_employer/contbr_occupation: this can definitely be useful
# contb_receipt_amt/contb_receipt_dt: info on donations
# type of form: SA17A (individual donation), SB28A (refund to individual), SA18 (transfer from authorized committee)
# election_tp: G2016 or P2016 (general/primary)
```
```{r Cities}
#just a model for getting data on a specific city
Newton = newjersey[contbr_city == 'NEWTON',]
```

```{r subsetting}
#subsetting to primary/general 2016 elections (don't want 2020 data)
newjersey = subset(newjersey, 
                     election_tp == 'G2016' | 
                       election_tp == 'P2016')
```

```{r candidates}
#function to rename candidates
candidate_rename <- function(candidate) {
  pos <- regexpr(',', candidate)[1] #find position of comma
  return(substr(candidate, 1, pos-1))
}

newjersey['cand_nm'] <- apply(newjersey['cand_nm'], 1, candidate_rename)

newjersey$cand_nm <- factor(newjersey$cand_nm)

#histogram of entries for each candidate faceted by general/primary election
ggplot(aes(x = cand_nm), 
       data = newjersey) +
  geom_histogram(stat = 'count') +
  facet_wrap(~election_tp)

#histogram of entries for each candidate colored according to general/primary election
ggplot(aes(x = cand_nm), 
       data = subset(newjersey, 
                     election_tp == 'G2016' | 
                       election_tp == 'P2016')) +
  geom_histogram(stat = 'count', aes(color = election_tp,
                                     fill = election_tp))

#bar graph of total contributions to each candidate
by_candidate <- group_by(newjersey, cand_nm)
contb_totals <- summarise(by_candidate,
                          total = sum(contb_receipt_amt),
                          n = n())

ggplot(aes(x = cand_nm, y = total), 
       data = contb_totals) +
  geom_bar(stat = 'identity') +
  scale_y_continuous(labels = comma)

#bar graph of total contributions to each candidate split out by general/primary elections
by_candidate_election <- group_by(newjersey, cand_nm, election_tp)
contb_totals_election <- summarise(by_candidate_election,
                                   total = sum(contb_receipt_amt),
                                   n = n())

ggplot(aes(x = cand_nm, y = total, fill = election_tp), 
       data = contb_totals_election) +
  geom_bar(stat = 'identity', position = position_dodge()) + 
  scale_y_continuous(labels = comma)

#examining median donation by candidate, labeling the total # of donations. only looking at primary donations. 
contb_stats <- summarise(by_candidate_election,
                         total = sum(contb_receipt_amt),
                         mean = mean(contb_receipt_amt),
                         median = mean(contb_receipt_amt),
                         n = n())

ggplot(aes(x = cand_nm, y = median, label = n, fill = log(n)), 
       data = subset(contb_stats, n > 10 & election_tp == 'P2016')) + #primaries only, add # of contributions above bars
  geom_bar(stat = 'identity', position = position_dodge()) + 
  scale_y_continuous(labels = comma) + 
  geom_text(aes(y = median + 100)) + 
  scale_fill_continuous(low = 'darkgoldenrod2', high = 'midnightblue') #color scale according to number of donations 

```

```{r new variables}
#candidate gender
assign_candidate_gender <- function(candidate) {
  gender <- 'M'
  if ((candidate == 'Clinton') | (candidate == 'Fiorina')) {
    gender <- 'F'
  } 
  return(gender)
}

newjersey$cand_gender <- factor(apply(newjersey['cand_nm'], 1,
                                      assign_candidate_gender))

#candidate party
assign_candidate_party <- function(candidate) {
  if (candidate == 'Johnson') {
    return('Libertarian')
  } else if (candidate == 'Stein') {
    return('Green')
  } else if (candidate == 'McMullin') {
    return('Independent')
  } else if (candidate %in% c('Clinton', 'Sanders', 'O\'Malley', 
                              'Webb', 'Lessig')) {
    return('Democratic')
  } else {
    return('Republican')
  }
}

newjersey$cand_party <- factor(apply(newjersey['cand_nm'], 1,
                                     assign_candidate_party))

#contributor gender
get_first_name <- function(contrib) {
  pos <- regexpr(',', contrib)[1] #find position of comma
  first <- substr(contrib, pos + 2, nchar(contrib))
  space <- regexpr(' ', first)[1] #in cases with space in name
  if (space > -1) {
    first <- substr(first, 1, space-1)
  }
  return(first)
}
  
assign_contrib_gender <- function(contrib) {
  year_min = 1932 #default
  year_max = 1998 #2016 - 18 years (nobody younger than this is giving money)
  name = get_first_name(toString(contrib))
  return(gender(name, years = c(year_min, year_max))$gender)
}

assign_contrib_gender_firstname <- function(contrib_firstname) {
  year_min = 1932 #default
  year_max = 1998 #2016 - 18 years (nobody younger than this is giving money)
  return(gender(contrib_firstname, years = c(year_min, year_max))$gender)
}

#newjersey$contbr_gender <- factor(apply(newjersey['contbr_nm'], 1, 
#                                        assign_contrib_gender))

```


```{r employers}
#creating subsets of data based on employers
universities <- c('RUTGERS UNIVERSITY', 'PRINCETON UNIVERSITY')
banks <- c('JPMORGAN CHASE', 'MORGAN STANLEY', 'BANK OF AMERICA', 
           'CITIGROUP')
pharma <- c('NOVARTIS', 'JOHNSON & JOHNSON', 'BAYER')
tech <- c('GOOGLE', 'FACEBOOK', 'MICROSOFT')

#only want to look at primaries for this study
primaries <- subset(newjersey, election_tp == 'P2016')

univ_donors <- subset(primaries, contbr_employer %in% universities)
bank_donors <- subset(primaries, contbr_employer %in% banks)
pharma_donors <- subset(primaries, contbr_employer %in% pharma)
tech_donors <- subset(primaries, contbr_employer %in% tech)

#looking at donors who are employed by the state's two 'largest' universities, Rutgers and Princeton. we would expect more of a liberal bent here
ggplot(aes(x = cand_nm), 
       data = univ_donors) +
  geom_histogram(stat = 'count')

#hahahahaahah

#now for banks - expect more Republican
ggplot(aes(x = cand_nm), 
       data = bank_donors) +
  geom_histogram(stat = 'count') +
  facet_wrap(~election_tp)

#slightly i guess, but not as stark - still predominantly for Clinton

#pharma also should, in theory, skew Republican. but perhaps not Trump
ggplot(aes(x = cand_nm), 
       data = pharma_donors) +
  geom_histogram(stat = 'count') +
  facet_wrap(~election_tp)

#not much to see here once again... two ideas. first, want to look at $$, instead of number of donations. second, perhaps this is just an employed vs. unemployed situation? 

by_candidate <- group_by(bank_donors, cand_nm)
contb_totals <- summarise(by_candidate,
                                   total = sum(contb_receipt_amt),
                                   n = n())

ggplot(aes(x = cand_nm, y = total), 
       data = contb_totals) +
  geom_bar(stat = 'identity', position = position_dodge()) + 
  scale_y_continuous(labels = comma)

#tested with pharma

#want to see proportion of total donations given to each candidate
by_candidate <- group_by(pharma_donors, cand_nm, election_tp)
contb_totals <- summarise(by_candidate,
                                   total = sum(contb_receipt_amt),
                                   n = n())

ggplot(aes(x = cand_nm, y = total/sum(total)), 
       data = contb_totals_election) +
  geom_bar(stat = 'identity', position = position_dodge()) + 
  scale_y_continuous(labels = comma)

#want to see proportion for each industry plotted together
add_proportion <- function(totals) {
  totals$proportion <- totals$total/sum(totals$total)
  return(totals)
}

#prep by-industry data
get_industry_totals <- function(industry_data, industry) {
  by_candidate <- group_by(industry_data, cand_nm)
  industry_totals <- summarise(by_candidate,
                            total = sum(contb_receipt_amt),
                            n = n(),
                            industry = industry)
  return(add_proportion(industry_totals))
}

univ_donors2 <- get_industry_totals(univ_donors, 'univ')
bank_donors2 <- get_industry_totals(bank_donors, 'bank')
pharma_donors2 <- get_industry_totals(pharma_donors, 'pharma')
tech_donors2 <- get_industry_totals(tech_donors, 'tech')

industries <- rbind(univ_donors2, bank_donors2, pharma_donors2, tech_donors2)

ggplot(aes(x = cand_nm, y = proportion, fill = industry),
       data = industries) + 
  geom_bar(stat = 'identity', position = position_dodge())

#what do we see here? i think there are a few conclusions we can draw. 
#1: the proportion of donations going to Clinton are *remarkably* consistent across the four industry grouips. in all cases it is between 65 and 75%. 
#2: workers at the big banks are outliers in two ways: first, they give a far higher proportion to the Republican candidates (particularly Bush and Christie - note Christie is probably a bit of a special case here since we are exmaining only New Jersey voters). second, they give a far *lower* proportion to Bernie Sanders; this makes perfect sense: there was no bigger enemy of the banks in the campaign than Sanders, who had the most consistently populist message of all of the significant candidates. 
#3: a couple of fun outliers worth noting: about 5% of tech money went to Larry Lessig, a thoroughly inconsequential candidate who likely drew most of his support from the technocratic class (he's a Harvard professor). also, John Kasich's only donations in this subset of the data came from tech workers! 

```

```{r cities}
by_city <- group_by(newjersey, contbr_city, cand_party)
city_totals <- summarise(by_city,
                                   total = sum(contb_receipt_amt),
                                   n = n())

#examining data:
republican_cities <- subset(city_totals, cand_party == 'Republican')
republican_cities <- republican_cities[order(-republican_cities$n),]
democratic_cities <- subset(city_totals, cand_party == 'Democratic')
democratic_cities <- democratic_cities[order(-democratic_cities$n),]

princeton <- subset(primaries, contbr_city == 'PRINCETON')
morristown <- subset(primaries, contbr_city == 'MORRISTOWN')
toms_river <- subset(primaries, contbr_city == 'TOMS RIVER')
bayonne <- subset(primaries, contbr_city == 'BAYONNE')

get_city_totals <- function(city_data, city) {
  by_candidate <- group_by(city_data, cand_nm)
  industry_totals <- summarise(by_candidate,
                            total = sum(contb_receipt_amt),
                            n = n(),
                            city = city)
  return(add_proportion(industry_totals))
}

princeton2 <- get_city_totals(princeton, 'Princeton')
morristown2 <- get_city_totals(morristown, 'Morristown')
toms_river2 <- get_city_totals(toms_river, 'Toms River')
bayonne2 <- get_city_totals(bayonne, 'Bayonne')

cities <- rbind(princeton2, morristown2, toms_river2, bayonne2)

ggplot(aes(x = cand_nm, y = proportion, fill = city),
       data = cities) + 
  geom_bar(stat = 'identity', position = position_dodge())

#for this study, we're examining contributions from four cities/towns in New Jersey, each representing a different archetype:
# - Princeton: a distincly politically active town, first in the state in terms of total money contributed and money contributed to Democrats
# - Morristown: had the highest $ amount of contributions to Republicans and a roughly equivalent amount given to Democrats
# - Bayonne: had a high number of low-$ amount donations
# - Toms River: by far the highest number of Republican contributions

#a few conclusions/observations:
#1: Bayonne, the city that had a disproportionately high number of small contributions, had a much higher % of its total going to Bernie Sanders, who was known throughout the primaries as the leading receiver of small contributions. for Republicans in Bayonne, almost all of the money went to Christie.
#2: Toms River was the only city that gave a notable percentage to Trump (about 15% of contributions); about half of the money went to Christie, with less than 10% going to Ms. Clinton or Mr. Sanders. 
#3: As noted earlier, Morristown has a pretty even split: Mr. Christie received nearly half of the money contributed, but Ms. Clinton received 35%. 
#4: A clear majority of Princeton money went to Ms. Clinton (about 55%), with Mr. Sanders receiving the second-most (about 15%). A few of the Republicans who were more popular among the "establishment", Jeb Bush and Marco Rubio, saw their best performance in this sample in Princeton. 

```

```{r gender}

##this is all example code to figure out how to do this process:
##---------------------------
all_contbrs <- data.frame(unique(contbr_nm)) #df of unique contbrs
colnames(all_contbrs) <- c('contbr_nm')
all_contbrs$first_name <- apply(all_contbrs['contbr_nm'], 1,
                                get_first_name) #only care about first names
some_contbrs <- data.frame(head(all_contbrs, 100))
ptm <- proc.time()
some_contbrs$contbr_gender <- apply(some_contbrs['first_name'], 1, 
                                    assign_contrib_gender_firstname)
proc.time() - ptm

#testing out hash functionality
h <- hash(some_contbrs$first_name, some_contbrs$contbr_gender)
h$LINDA
h[['LINDA']]

get_gender <- function(name, hash) {
  return(hash[[name]])
}

some_contbrs$contbr_gender2 <- apply(some_contbrs['first_name'], 1, 
                                     get_gender, hash = h)
##----------------------------------------------------------

#first get unique first names:
all_contbrs <- data.frame(unique(contbr_nm)) #df of unique contbrs
colnames(all_contbrs) <- c('contbr_nm')
all_contbrs$first_name <- apply(all_contbrs['contbr_nm'], 1,
                                get_first_name)
all_firsts <- data.frame(unique(all_contbrs$first_name))
colnames(all_firsts) <- c('first_name')

#now run get gender on all of them (this will take ~30-45 minutes for ~5000 names)
all_firsts$gender <- apply(all_firsts['first_name'], 1, 
                                    assign_contrib_gender_firstname)
all_firsts2 <- all_firsts[-c(3178),]
gender_hash <- hash(all_firsts2$first_name, all_firsts2$gender)

#now hash to assign to master df
newjersey$first_name <- apply(newjersey['contbr_nm'], 1,
                              get_first_name)
#need to cut out the entries wihtout a first name (there are only a few)
newjersey <- newjersey[newjersey['first_name']!='',]
newjersey$contbr_gender <- apply(newjersey['first_name'], 1, 
                                 get_gender, hash = gender_hash)
newjersey$contbr_gender <- factor(newjersey$contbr_gender)

##testing --------
subnewjersey <- head(newjersey, 47600)
subnewjersey$contbr_gender <- apply(subnewjersey['first_name'], 1,
                                    get_gender, hash = gender_hash)

subnewjersey <- newjersey[c(47550:47600),]$first_name
names <- newjersey[newjersey['first_name']!='',]

```

```{r gender2}
primaries <- subset(newjersey, election_tp == 'P2016')
by_candidate <- group_by(primaries, cand_nm, contbr_gender)
contb_totals_election <- summarise(by_candidate,
                                   total = sum(contb_receipt_amt),
                                   n = n())

ggplot(aes(x = cand_nm, y = total, fill = contbr_gender), 
       data = contb_totals_election) +
  geom_bar(stat = 'identity', position = position_dodge()) + 
  scale_y_continuous(labels = comma)

```

```{r income}
#read in income data and subset to newjersey
income <- read.csv('zipcode2014/14zpallnoagi.csv')
njincome <- subset(income, STATE == 'NJ')
#read in zipcode data
zipcodes <- read.csv('censuszipcode.csv')
colnames(zipcodes) <- c('zip', 'population')

#make hash of zipcode data
ziphash <- hash(zipcodes$zip, zipcodes$population)
get_population <- function(zip, hash) {
  return(hash[[toString(zip)]])
}

njincome$population <- apply(njincome['ZIPCODE'], 1, get_population,
                           hash = ziphash)
njincome <- subset(njincome, population != 'NULL')
njincome$population <- unlist(njincome$population)
#njincome$population <- factor(njincome$population) shouldn't factor?

#get AGI per capita
njincome$agiPC <- as.numeric(njincome$A00100)/as.numeric(njincome$population)

#clean up zip codes in contribution data
clean_zip <- function(zip) {
  zip <- toString(zip)
  if (nchar(zip) == 0) { #for blank rows
    return(zip)
  }
  if (nchar(zip) == 9) { #zip+4 w/ non-0 leading digit
    return(as.numeric(substr(zip, 1, 5)))
  } else if (nchar(zip) == 8) { #zip+4 w/ 0 as leading digit
    return(as.numeric(substr(zip, 1, 4)))
  } else { #other stuff
    return(as.numeric(zip))
  }
}

#create cleaned_zips
newjersey$cleaned_zips <- factor(apply(newjersey['contbr_zip'], 1,
                                       clean_zip))

slimnjincome <- njincome[c('ZIPCODE','A00100','population',
                           'agiPC')]

njmerged <- merge(newjersey, slimnjincome, 
                      by.x = 'cleaned_zips', by.y = 'ZIPCODE')
#want to defactor these:
njmerged$population <- as.numeric(as.character(njmerged$population))
njmerged$agiPC <- as.numeric(as.character(njmerged$agiPC))

```

```{r income analysis}
#simple plot showing roughly the distro of donations in zips of each income
ggplot(aes(x = agiPC), 
       data = njmerged) +
  geom_histogram(binwidth = 200) +
  coord_cartesian(xlim = c(0, 15000))

#adding in split by party
ggplot(aes(x = agiPC), 
       data = njmerged) +
  geom_histogram(binwidth = 200, aes(fill = cand_party)) +
  coord_cartesian(xlim = c(0, 15000))

#normal plots subsetting by lower income zip codes and higher income zips
by_candidate_low <- group_by(subset(njmerged, agiPC < 2000), cand_nm)
contb_totals <- summarise(by_candidate_low,
                          total = sum(contb_receipt_amt),
                          n = n())

ggplot(aes(x = cand_nm, y = total),
       data = contb_totals) + 
  geom_bar(stat = 'identity', position = position_dodge())

##i want to see where the proportion of moeny in zips 0-10% and 90-100% is going (obv total will be higher in 90-100%). 
#get cutoffs:
quantile(njmerged$agiPC, prob = seq(0, 1, length = 11), type = 5)
#will use roughly $1500 and $5000 to show 'low income' and 'high income'
njmerged$lowincome <- (njmerged$agiPC <= 1500)
njmerged$highincome <- (njmerged$agiPC >= 5000)
njmerged$incomerank <- (2 - njmerged$lowincome + njmerged$highincome)
njmerged$incomerank <- factor(njmerged$incomerank)

by_candidate_income <- group_by(subset(njmerged, election_tp == 'P2016'), cand_nm, incomerank)
income_totals <- summarise(by_candidate_income, 
                           total = sum(contb_receipt_amt),
                           n = n())

income_split <- function(row) {
  total <- as.numeric(row[['total']])
  rank <- as.numeric(row[['incomerank']])
  return(total/sum(subset(income_totals, incomerank == rank)$total))
}

income_totals$proportion <- apply(income_totals, 1, income_split)

#SUCCESS
ggplot(aes(x = cand_nm, y = proportion, fill = incomerank),
       data = income_totals) + 
  geom_bar(stat = 'identity', position = position_dodge())

#this is perfect. takeaways:
#Sanders by far most popular among lowest
#Clinton support rather consistent
#Christie and Bush are men of the rich
#Trump also working class

#now going to try the same, but with five even quintiles instead
quantile(njmerged$agiPC, prob = seq(0, 1, length = 11), type = 5)

njmerged$incomerank2 <- (njmerged$agiPC <= 1785.95)
njmerged$incomerank2 <- njmerged$incomerank2 + (njmerged$agiPC <= 2528.81)
njmerged$incomerank2 <- njmerged$incomerank2 + (njmerged$agiPC <= 3416.53)
njmerged$incomerank2 <- njmerged$incomerank2 + (njmerged$agiPC <= 4566.53)
njmerged$incomerank2 <- 5 - njmerged$incomerank2

njmerged$incomerank2 <- factor(njmerged$incomerank2)

by_candidate_income <- group_by(subset(njmerged, election_tp == 'P2016'), cand_nm, incomerank2)
income_totals <- summarise(by_candidate_income, 
                           total = sum(contb_receipt_amt),
                           n = n())

income_split <- function(row) {
  total <- as.numeric(row[['total']])
  rank <- as.numeric(row[['incomerank2']])
  return(total/sum(subset(income_totals, incomerank2 == rank)$total))
}

income_totals$proportion <- apply(income_totals, 1, income_split)

ggplot(aes(x = cand_nm, y = proportion, fill = incomerank2),
       data = income_totals) + 
  geom_bar(stat = 'identity', position = position_dodge()) +
  scale_fill_brewer(palette = 'greens') + 
  theme(panel.background = element_rect(fill = 'rosybrown1'))

```

```{r more gender stuff}
primaries <- subset(njmerged, election_tp == 'P2016')
by_candidate <- group_by(primaries, cand_nm, contbr_gender)
contb_totals_election <- summarise(by_candidate,
                                   total = sum(contb_receipt_amt),
                                   n = n())

ggplot(aes(x = cand_nm, y = total, fill = contbr_gender), 
       data = contb_totals_election) +
  geom_bar(stat = 'identity', position = position_dodge()) + 
  scale_y_continuous(labels = comma)

#this shows total giving by gender. it isn't super useful, because it's clear that the women in the sample have given less overall. but we can see that women have given more to clinton, but are "outdone" by men in all of the others

#skipping right to the end, let's look at republican candidates only, and proportions by gender
by_republican_candidate_gender <- group_by(subset(primaries, cand_party == 'Republican'), cand_nm, contbr_gender)
comb_totals_republican <- summarise(by_republican_candidate_gender,
                                   total = sum(contb_receipt_amt),
                                   n = n())

republican_split <- function(row) {
  total <- as.numeric(row[['total']])
  gender <- row[['contbr_gender']]
  return(total/sum(subset(comb_totals_republican,contbr_gender == gender)$total))
}

comb_totals_republican$proportion <- apply(comb_totals_republican, 1, republican_split)

ggplot(aes(x = cand_nm, y = proportion, fill = contbr_gender), 
       data = comb_totals_republican) +
  geom_bar(stat = 'identity', position = position_dodge()) + 
  scale_y_continuous(labels = comma)

#assuming this is correct (i feel like i have to double-check) this is pretty astounding! there is barely any difference at all! no republican division by sex! wow!

#crazy. it's real! 
ggplot(aes(x = cand_nm, y = total, fill = contbr_gender), 
       data = comb_totals_republican) +
  geom_bar(stat = 'identity', position = position_dodge()) + 
  scale_y_continuous(labels = comma)
```

```{r dates}
summary(njmerged$contb_receipt_dt)
ggplot(aes(x = contb_receipt_dt, y = contb_receipt_amt),
       data = njmerged) +
  geom_bar(stat = 'identity')

by_date <- group_by(njmerged, contb_receipt_dt)
date_totals <- summarise(by_date,
                         total = sum(contb_receipt_amt),
                         n = n())
ggplot(aes(x = contb_receipt_dt, y = total),
       data = date_totals) + 
  geom_bar(stat = 'identity') + 
  scale_x_date(date_breaks = '1 month') #scale x date only works with Dates

#this works
njmerged$newdates <- as.Date(njmerged$contb_receipt_dt,
                             format = '%d-%b-%y')
by_date <- group_by(njmerged, newdates)
date_totals <- summarise(by_date,
                         total = sum(contb_receipt_amt),
                         n = n())
ggplot(aes(x = newdates, y = total),
       data = date_totals) + 
  geom_bar(stat = 'identity') +
  scale_x_date(date_breaks = '1 month',
               date_labels = '%b %y')

#that graph roughly shows donations over time

```

```{r switching}
#want to look at people who gave to Republicans in the primary and then to Hillary in the general. also people who gave to Sanders in the primary and then Trump in the general.

#donors to Sanders in the primary:
berniebros <- subset(njmerged, cand_nm == 'Sanders')
berniebrosgen <- subset(njmerged, contbr_nm %in% berniebros$contbr_nm)
berniebrosgen <- subset(berniebrosgen, election_tp == 'G2016')
berniebrossum <- group_by(berniebrosgen, cand_nm)
berniegeneral <- summarise(berniebrossum, 
                           total = sum(contb_receipt_amt),
                           n = n())

ggplot(aes(x = cand_nm, y = total),
        data = berniegeneral) +
  geom_bar(stat = 'identity') +
  scale_y_continuous(labels = comma)

#basically everyone from Bernie gives to Hillary

#donors to non-trump Republicans in primary
nevertrumpers <- subset(subset(njmerged, election_tp == 'P2016'), (cand_party == 'Republican' & cand_nm != 'Trump'))
nevertrumpersgen <- subset(njmerged, contbr_nm %in% nevertrumpers$contbr_nm)
nevertrumpersgen <- subset(nevertrumpersgen, election_tp == 'G2016')
nevertrumperssum <- group_by(nevertrumpersgen, cand_nm)
nevertrumpersgeneral <- summarise(nevertrumperssum,
                                  total = sum(contb_receipt_amt),
                                  n = n())

ggplot(aes(x = cand_nm, y = total),
       data = nevertrumpersgeneral) +
  geom_bar(stat = 'identity')

#confusing b/c it shows a lot of donations to non-trump republicans, not sure how that's possible. perhaps i should subset to the point where trump was officially nominated?

nevertrumpersgen <- subset(nevertrumpersgen, newdates >= '2016-05-04')
nevertrumperssum <- group_by(nevertrumpersgen, cand_nm)
nevertrumpersgeneral <- summarise(nevertrumperssum,
                                  total = sum(contb_receipt_amt),
                                  n = n())
ggplot(aes(x = cand_nm, y = total),
       data = nevertrumpersgeneral) +
  geom_bar(stat = 'identity')

#this is better; now shows 4 candidates (no Stein) + refunds to Rubio and Cruz, which i will eliminate in the bigger dataset later. there is a solid amount of Clinton cash! more than 50% of trump total

#now let's look back and see who the Clinton donors were giving to in the primary
republicantoclinton <- subset(nevertrumpersgen, cand_nm == 'Clinton')
rtcprimaries <- subset(njmerged, election_tp == 'P2016' &
                         contbr_nm %in% republicantoclinton$contbr_nm)
rtcgrouped <- group_by(rtcprimaries, cand_nm)
rtcsums <- summarise(rtcgrouped,
                     total = sum(contb_receipt_amt),
                     n = n())

ggplot(aes(x = cand_nm, y = total),
       data = rtcsums) +
  geom_bar(stat = 'identity')

#there is a lot of Clinton in here; maybe people started to give to Clinton earlier? maybe i need a stronger unique identifier for donors? (there could be duplicate names)

```


# Univariate Plots Section
```{r echo=FALSE, Univariate_Plots}


```

# Univariate Analysis

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?



# Bivariate Plots Section
```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

### What was the strongest relationship you found?




# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------
```{r sources}

#zipcode: https://blog.splitwise.com/2013/09/18/the-2010-us-census-population-by-zip-code-totally-free/
#income: https://www.irs.gov/uac/soi-tax-stats-individual-income-tax-statistics-zip-code-data-soi


```

# Reflection
